Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron

### 2022-08-10 Rmarkdown/paper-si.Rmd
edits for submission to chemrxiv

This document contains X tables and X figures.

@@

```{r packages, message=FALSE}
library("tidyverse")
library("fffprocessr")
source(here::here("R/functions.R"))
```

@@

param_names <- tibble::tribble(
   ~param,                       ~plot_param,
  "208Pb",       "Pb (&mu;g L<sup>-1</sup>)",
    "31P",        "P (&mu;g L<sup>-1</sup>)",
   "57Fe",       "Fe (&mu;g L<sup>-1</sup>)",
   "LS90", "90&deg; light scattering signal",
"UV254_2",                "UV<sub>254</sub>",
   "ls90",         "Radius of gyration (nm)"
)

@@

```{r fig-s1, fig.cap="**Figure S1:** Fractograms representing pure solutions of sodium hexameta- and trimeta-phosphate.", fig.height = 2.5, fig.width = 3.33}

@@

```{r mals-plot, eval=FALSE}

mals_data %>% 
  filter(time > 10, date < "2021-12-01") %>% 
  ggplot(aes(time, conc, col = sample, group = interaction(date, sample))) + 
  facet_wrap(
    vars(fct_reorder(param, str_extract(param, "\\d+") %>% as.numeric())), 
    scales = "free_y"
  ) + 
  geom_line() + 
  theme(legend.position = "bottom") +
  guides(col = guide_legend(nrow = 2)) +
  labs(col = NULL)
```

>>>>

This document contains 1 table and 3 figures.

@@

```{r packages, message=FALSE}
library("tidyverse")
library("fffprocessr")
library("ggtext")
source(here::here("R/functions.R"))
```

@@

@@

```{r mals-plot, fig.cap="**Figure S1.** Baseline-corrected multi-angle light scattering data (triplicate sodium hexametaphosphate suspensions)."}

param_mals <- distinct(mals_data, param)$param
striplabs <- str_replace(param_mals, "(ls)(\\d+)", "\\2&deg;")
names(striplabs) <- param_mals
striplabs <- as_labeller(striplabs)

plot_mals <- function(suspension) {
  mals_data %>% 
    filter(time > 30, sample == suspension) %>% 
    ggplot(aes(time, conc, group = interaction(date, sample))) + 
    facet_wrap(
      vars(fct_reorder(param, as.numeric(str_extract(param, "\\d+")))),
      labeller = striplabs,
      # scales = "free_y"
    ) + 
    geom_line(size = .3) + 
    theme(
      legend.position = "bottom",
      strip.text = ggtext::element_markdown(),
      axis.text = element_text(size = 5)
    ) +
    labs(
      x = "Retention time (min)",
      y = "Scattering intensity above baseline",
      col = NULL
    )

}

plot_mals("3PP")

```{r mals-plot-2, fig.cap="**Figure S2.** Baseline-corrected multi-angle light scattering data (triplicate sodium silicate suspensions)."}
plot_mals("100SiO2")
```

```{r fig-s1, fig.cap="**Figure S3:** Fractograms representing pure solutions of sodium hexameta- and trimeta-phosphate.", fig.height = 2, fig.width = 3.33}

@@

### 2022-08-10 Rmarkdown/paper.Rmd
edits for submission to chemrxiv

Less is known about how polyphosphates and silicates impact iron release from corrosion scale. In the short term, an elevated dose of sodium silicate may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available. The effect of polyphosphate on iron colloid release from corrosion scale is also unclear, but it could be important: iron colloids are important vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016]. Moreover, colloidal lead can be difficult to remove at the point of use [@lytle_lead_2020; @pan_lead_2021; @purchase_understanding_2020].

@@

@@

The deviation of the channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients drawn from literature [@clough_role_1981; @gaigalas_diffusion_1992]. The FFF method for the globular protein runs is summarized in Table S1.

@@

where $K$ is a constant, $c$ and $M$ are the concentration and molar mass of the analyte, respectively, $\lambda$ is the wavelength of the incident light, $\theta$ is the scattering angle, $\eta_0$ is the index of refraction of the solvent, and $R$ is the Rayleigh ratio, or the scattering intensity above baseline. For environmental colloids, $K$, $c$, and $M$ are usually unknown, but $r_g^2$ can be estimated by regressing $R^{-1}$ on $sin^2(\theta/2)$. That is,

@@

```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999.", message=FALSE}

@@

Sodium silicate and hexametaphosphate suspensions yielded similar iron fractograms (Figure \@ref(fig:figure-fff)), with primary and secondary light scattering peaks. These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes. Iron was also present in a broad peak, eluting between approximately 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm. This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017]. In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S1).

@@

```{r table-radii}
radii %>% 
  arrange(sample, rep, peak) %>% 
  left_join(pcent_tot, by = c("sample", "peak", "rep")) %>% 
  transmute(
    Suspension = fct_recode(sample, "Sodium silicate" = "100SiO2", "Hexametaphosphate" = "3PP") %>% 
      unfill_vec("-"),
    Replicate = unfill_vec(rep, "-"),
    Peak = peak,
    `Retention time (min)` = signif(time, 3), 
    `Radius of gyration (nm)` = signif(rg_zimm, 3),
    `Spherical equivalent geometric diameter (nm)` = signif(d_geom, 3),
    `% total Fe` = round(pcent, 1)
  ) %>% 
  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters.")
```

@@

We integrated the ^31^P and ^57^Fe signals over each fractogram---excluding the focus period---to estimate the concentrations of each analyte. Then, we compared these estimates to the concentrations determined by direct ICP-MS quantification (i.e., no FFF). Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe` in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)). While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions. And since recovery of P from a sodium hexametaphosphate standard was 111%, the partial P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1). The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane. In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].

>>>>

Less is known about how polyphosphates and silicates impact iron release from corrosion scale. In the short term, an elevated dose of sodium silicate may promote colloidal iron release from corroded iron water mains [@li_impact_2020], but little characterization of these colloids is available. The effect of polyphosphate on iron colloid release from corrosion scale is also unclear, but it could be important: iron colloids are vectors for trace element transport in drinking water systems, which is particularly relevant for controlling lead in drinking water [@trueman_understanding_2016; @trueman_characterizing_2019; @trueman_new_2016]. Moreover, colloidal lead can be difficult to remove at the point of use [@lytle_lead_2020; @pan_lead_2021; @purchase_understanding_2020].

@@

The deviation of the FFF channel thickness from its nominal value of 500 µm was estimated using a method described elsewhere [@litzen_separation_1993], using ferritin and bovine serum albumin as globular protein standards with diffusion coefficients drawn from literature [@clough_role_1981; @gaigalas_diffusion_1992]. The FFF method for the globular protein runs is summarized in Table S1. The MALS detector was calibrated and normalized using bovine serum albumin, polystyrene sulfonate sodium salt (63.9 kDa), and a mixture of latex particles with nominal geometric diameters of 60, 125, 350 nm.

@@

@@

where $K$ is a constant, $c$ and $M$ are the concentration and molar mass of the analyte, respectively, $\lambda$ is the wavelength of the incident light, $\theta$ is the scattering angle, $\eta_0$ is the index of refraction of the solvent, and $R$ is the Rayleigh ratio, or the scattering intensity above baseline at each detector angle (Figures S1--2). For environmental colloids, $K$, $c$, and $M$ are usually unknown, but $r_g^2$ can be estimated by regressing $R^{-1}$ on $sin^2(\theta/2)$. That is,

@@

```{r figure-zimm, fig.cap="The Zimm model described the variation in light scattering over the selected range of scattering angles, with r^2^ values of 0.983--0.999. Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.", message=FALSE}

@@

Sodium silicate and hexametaphosphate suspensions yielded similar iron fractograms (Figure \@ref(fig:figure-fff)), with primary and secondary light scattering peaks. These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes. Iron was also present in a broad peak, eluting between approximately 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm. This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017]. In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at `r paste(round(range(peak_pos_p$time), 1), collapse = "--")` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).

@@

```{r table-radii}
radii %>% 
  arrange(sample, rep, peak) %>% 
  left_join(pcent_tot, by = c("sample", "peak", "rep")) %>% 
  transmute(
    Suspension = fct_recode(sample, "Sodium silicate" = "100SiO2", "Hexametaphosphate" = "3PP") %>% 
      unfill_vec("-"),
    Replicate = unfill_vec(rep, "-"),
    Peak = peak,
    `Retention time (min)` = signif(time, 3), 
    `Radius of gyration (nm)` = signif(rg_zimm, 3),
    `Spherical equivalent geometric diameter (nm)` = signif(d_geom, 3),
    `% total Fe` = round(pcent, 1)
  ) %>% 
  knitr::kable(caption = "Estimated radii of gyration and calculated geometric (equivalent spherical) diameters. Peaks 1 and 2 eluted at retention times of 34--35 and 44--45 minutes.")
```

@@

We integrated the ^31^P and ^57^Fe signals over each fractogram---excluding the focus period---to estimate the concentrations of each analyte. Then, we compared these estimates to the concentrations determined by direct ICP-MS quantification (i.e., no FFF). Recovery of ^57^Fe was `r recovery_text$percent_recovery_100Si_57Fe` and `r recovery_text$percent_recovery_3PP_57Fe`% in the sodium silicate and hexametaphosphate suspensions (Table \@ref(tab:table-recov)). While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions. And since recovery of P from a sodium hexametaphosphate standard was 111%, the partial P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1). The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane. In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].
